The increasing cost of happiness

### 2021-06-06 src/draft-SSM1.Rmd
Added log-linear models. Closes issue #10

Subjective wellbeing is not a unitary entity [@diener2017findings]; studies typically distinguish between life satisfaction, the cognitive appraisal of one’s own accomplishments, and affective wellbeing, one’s prevailing affective state, emotional mood, or everyday experience of happiness [e.g., @kettlewell2020differential]. Kahneman & Deaton [-@kahneman2010high] showed that affective wellbeing increased with household income up to a point (USD75,000), but after that, increasing income had little further effect. Conversely cognitive wellbeing continued to increase with income beyond USD75,000. Indeed, the difference between the two questions: “How satisfied are you with your life?” and “How happy are you these days?” has been identified as a crucial mediating factor in a meta-analysis of 111 studies on income and wellbeing [@howell2008relation; also @veenhoven2006rising]. Results since then have also shown that while both affective and cognitive wellbeing increase with income, the affective returns are decreasing [@killingsworth2021experienced; @sacks2012new].

Fundamentally, the existence of a change point in the relationship between income and affective wellbeing reveals an unacknowledged source of inequality in the distribution of wellbeing (i.e., happiness) in the economy. For instance, a change point of USD75,000 in 2008 was substantially more than the US median income of USD52,000 in the same year, indicating that the affective wellbeing of the poorest majority of the US population was tied to marginal changes in income while the affective wellbeing of a richer minority was not. Thus the change point represents the dollar value up to which income drives inequities in the distribution of affective wellbeing, such that a lower valued change point represents a more equitable distribution of affective wellbeing in the economy. Inequities in the distribution of wellbeing are increasingly relevant to governments and policy-makers due to the growing recognition that increasing income does not necessarily lead to equal changes in wellbeing [@clark2018four; @frijters2020happy]. Even prior to COVID-19, the World Gallup Poll has observed that affective wellbeing has decreased over the past decade in western Europe, North America, Australia and New Zealand, despite increases in income in the same countries [@sachs2019world]. However to date there has been no investigation of whether the relationship between income and affective wellbeing has changed over time which may contribute to these trends. In particular has the change point between income and affective wellbeing, and therefore the distribution of happiness between rich and poor, become more or less equitable in the last few decades?

@@

We used household economic panel data from Australia to provide the first investigation of whether changes in income and wellbeing have shifted the change point between 2001 to 2019. HILDA (the Household, Income, and Labour Dynamics in Australia survey) provides a representative sample of households in Australia with detailed measurements of income and subjective wellbeing each year. We distinguished between cognitive- and affective wellbeing as different components of subjective wellbeing, and evaluated how each varied with household income over time using piecewise linear regression to determine the change points in each year.

@@

We modelled the relationship between income and each wellbeing variable (cognitive and affective) using a log-linear model and a piecewise-linear model with a single change-point as a free parameter estimated from the data. The piecewise-linear model was chosen as the simplest extension of a linear model which can identify a change point (inflection) in the relationship between wellbeing and income. For modelling, income values were log transformed and then both measures of wellbeing and income were rescaled with a mean of zero and a SD of 1 (z-scores) for each year.

_Model Design_  
We adopted a Bayesian approach for estimating the log-linear and piecewise-linear model in the software Stan [@burkner2017; @rstan2019]. In each case the log-linear model was estimated as:

@@

_Model Selection_  
We compared the out-of-sample deviance of the log-linear and piecewise-linear model fits using the [Widely Applicable Information Criterion (WAIC)](https://bookdown.org/ajkurz/Statistical_Rethinking_recoded/overfitting-regularization-and-information-criteria.html#the-problem-with-parameters){target="_blank"}. The WAIC represents an approximation of the out-of-sample deviance that converges to the cross-validation accuracy in a large sample, with a penalty for the effective number of parameters (degrees of freedom). Thus using the out-of-sample deviance for model selection in combination with regularizing priors in our model design is a dual strategy to reduce overfitting and penalize overfitting. As with other deviance metrics, smaller negative WAIC values are better (i.e., indicate more accuracy).

@@

The nonlinear relationship between affective wellbeing and income (Figure 1, 2nd row) was consistently and negatively inflected (affective wellbeing increased less with income after the change point). By contrast the nonlinear relationship between cognitive wellbeing and income shown in the 4th row exhibited negative inflection (2001, 2015, 2019), positive inflection (2005), and no apparent inflection (2010).

The evidence from model selection revealed the nonlinear (piecewise) fit of _affective_-, but not _cognitive_ wellbeing, was superior to a linear fit in each year (Supplementary Materials, Figure S1). Overall the model comparison suggested that cognitive- and affective wellbeing have distinct relationships with household income; cognitive wellbeing tends to increase linearly with income, while a change point exists in the relationship between affective wellbeing and household income.

>>>>

Subjective wellbeing is not a unitary entity [@diener2017findings]; studies typically distinguish between life satisfaction, the cognitive appraisal of one’s own accomplishments, and affective wellbeing, one’s prevailing affective state, emotional mood, or everyday experience of happiness [e.g., @kettlewell2020differential]. Kahneman & Deaton [-@kahneman2010high] showed that affective wellbeing increased with household income up to a point (USD75,000), but after that, increasing income had little further effect. Conversely cognitive wellbeing continued to increase with income beyond USD75,000. Indeed, the difference between the two questions: “How satisfied are you with your life?” and “How happy are you these days?” has been identified as a crucial mediating factor in a meta-analysis of 111 studies on income and wellbeing [@howell2008relation; also @veenhoven2006rising]. Studies since then have also shown that while both affective and cognitive wellbeing increase with income, the affective returns are a decreasing function of income [@killingsworth2021experienced; @sacks2012new].

While studies agree that cognitive wellbeing tends to increase linearly, or log-linearly, with income [e.g., @stevenson2013subjective; but see @jebb2018happiness], the exact form of the functional relationship between affective wellbeing and income is contested: some studies show the presence of a changepoint after which increasing income has little further effect [@kahneman2010high; @jebb2018happiness]; other studies have found that affective wellbeing is a constant (albiet decreasing) log-linear function of income [@killingsworth2021experienced; @sacks2012newl]; and indeed at least one recent study has indicated a constant linear relationship [@twenge2020expanding]. Few studies have formally tested the functional form of the relationship, relying instead upon visual comparison after binning the data over income deciles or quintiles. In this regard, arbitrary decisions about bin width may have salient effects on the functional form, which in turn will influence the inferences drawn. For this reason, a formal comparison between linear, piecewise-linear and log-linear functions in the relationship between affective wellbeing and income (unbinned) is warranted. More fundementally, the existence of a change point in the relationship between income and affective wellbeing would reveal an unacknowledged source of inequality in the distribution of wellbeing (i.e., happiness) in the economy. For instance, a change point of USD75,000 in 2008 was substantially more than the US median income of USD52,000 in the same year, indicating that the affective wellbeing of the poorest majority of the US population was tied to marginal changes in income while the affective wellbeing of a richer minority was not. Thus a change point would represent the dollar value up to which income drives inequities in the distribution of affective wellbeing, such that a lower valued change point represents a more equitable distribution of affective wellbeing in the economy.

Inequities in the distribution of wellbeing are increasingly relevant to governments and policy-makers due to the growing recognition that increasing income does not necessarily lead to equal changes in wellbeing [@clark2018four; @frijters2020happy]. Even prior to COVID-19, the World Gallup Poll has observed that affective wellbeing has decreased over the past decade in western Europe, North America, Australia and New Zealand, despite increases in income in the same countries [@sachs2019world]. However to date there has been little investigation of whether the relationship between income and affective wellbeing has changed over time, which may contribute to these trends. In particular has the functional relationship between income and affective wellbeing, and therefore the distribution of happiness between rich and poor, become more or less equitable in the last few decades?

@@

We used household economic panel data from Australia to provide the first investigation of whether changes in the functional form between income and wellbeing have shifted between 2001 to 2019. HILDA (the Household, Income, and Labour Dynamics in Australia survey) provides a representative sample of households in Australia with detailed measurements of income and subjective wellbeing each year. We distinguished between cognitive- and affective wellbeing as different components of subjective wellbeing, and evaluated how each varied with household income over time using linear, log-linear and piecewise linear regression in each year.

@@

We modelled the relationship between income and each wellbeing variable (cognitive and affective) using a linear model, log-linear model, and a piecewise-linear model with a single change-point as a free parameter estimated from the data. The piecewise-linear model was chosen as the simplest extension of a linear model which can identify a change point (inflection) in the relationship between wellbeing and income. For modelling, both measures of wellbeing and income were rescaled with a mean of zero and a SD of 1 (z-scores) for each year. As each year was modelled separately this does not affect the results. Importantly, the coefficient for the change point in each year was returned to the original scale and transformed to real 2019 dollars when reported.

_Model Design_  
We adopted a Bayesian approach for estimating the linear, log-linear and piecewise-linear model in the software Stan [@burkner2017; @rstan2019]. The log-linear model was estimated after log transforming the income values, and the linear and log-linear model were then both estimated as:

@@

_Model Selection_  
We compared the out-of-sample deviance of the log-linear and piecewise-linear model fits using the [Widely Applicable Information Criterion (WAIC)](https://bookdown.org/ajkurz/Statistical_Rethinking_recoded/overfitting-regularization-and-information-criteria.html#the-problem-with-parameters){target="_blank"}. The WAIC represents an approximation of the out-of-sample deviance that converges to the cross-validation accuracy in a large sample, with a penalty for the effective number of parameters (degrees of freedom). Thus using the out-of-sample deviance for model comparison in combination with regularizing priors in our model design is a dual strategy to reduce overfitting and penalize overfitting. As with other deviance metrics, smaller WAIC values are better (i.e., indicate more accuracy).

@@

The piecewise relationship between affective wellbeing and income (Figure 1, 2nd row) was consistently and negatively inflected (affective wellbeing increased less with income after the change point). By contrast the piecewise relationship between cognitive wellbeing and income shown in the 4th row exhibited negative inflection (2001, 2015, 2019), positive inflection (2005), and no apparent inflection (2010).

The evidence from model selection revealed the piecewise-linear fit of _affective_-, but not _cognitive_ wellbeing, was superior to a log-linear or linear fit in each year (Supplementary Materials, Figure S1). Overall the model comparison suggested that cognitive- and affective wellbeing have distinct relationships with household income; cognitive wellbeing tends to increase linearly with income, while a change point exists in the relationship between affective wellbeing and household income.
